---
title: "Individual-based Integral Projection Models: The role of size-structure on extinction risk and establishment success"
author: "Sebastian Schreiber and Noam Ross"
date: "April 8, 2015"
output: html_document
---
$\def\N{\mathbb N}$
$\def\R{\mathbb R}$
$\def\P{\mathbb P}$
$\def\N{\mathbb N}$
$\def\S{\mathcal S}$
$\def\n{\mathbf n}$
$\def\s{\mathbb s}$
$\def\x{\mathbf x}$
$\def\y{\mathbf y}$
$\def\o{\%*\%}$

##Introduction

The likelihoods of extinction or establishment success lie on opposing sides of a theoretican's coin that has and continues to play an important role in questions in conservation biology, restoration ecology, biological invasions and population genetics. These likelihoods stem from populations consisting of a finite number of individuals each of which faces a positive risk of mortality on any given day. Moreover, these extinction and establishment likelihoods are shaped by population structure in which individuals may differ in age, size, geographical location, or other important distinguishing features. Multi-type branching processes, in which individuals survive, grow, and repoduce independent of one another, are a natural stochastic counterpart of matrix models (Caswell 2001). These models allow one to examine how extinction risk or establishment success depend on the population size and its distribution across a finite number of states (e.g. ages, size classes, patches). 

When collecting demographic data, one commonly makes continuous measurements (e.g. mass, length) about individuals. Consequently, since their introduction in the early 2000s, there has been a surge of interest in integral projection models (IPMs) that account for continuous population structure and correspond to infinite-dimensional analog of matrix models. The individual based counterparts of these models were effectively introduced in the 1950s by Theodore Harris. He called them continuous-state branching processes. Here, we describe a generalization of these models as well an explicit analytical method for computing time-dependent extinction probabilities associated with these individual based IPMs. We illustrate how to implement these methods in R using the IPM of XXX developed by XXX. Explicit sensitivity formulas for these extinction probabilities are also provided. 



##Models and methods

### The General Individual Based IPM
We consider an individual-based model where the set of all possible individual states (e.g. age, size, etc.) lies in a compact metric space $X$. For example, for the standard size-structured IPM, $X=[a,b]$ is the interval of sizes. For a mixture of age and size stucture, $X=\{1,\dots,T\}\times [a,b]$ where $T$ is the maximal age. As we only consider populations with a finite number of individuals, the state of the population at any point of the time is characterized by the sizes $(x_1,x_2,\dots, x_k)$  of individuals within the population  and how many individuals are of a given size $(n_1,n_2,\dots, n_k)$. Specifically, if there $n_1$ individuals of size $x_1$, $n_2$ individuals of size $x_2$,..., $n_k$ individuals of size $x_k$, then the state of the population is given by $(n_1,n_2,\dots, n_k; x_1,x_2,\dots, x_k)$. Following Harris (XXXX), we denote the set of all possible population states is
\[
\S= \{(\n,\x)=(n_1,\dots, n_k, x_1\dots , x_k): k,n_i\in \N, x_i \in X\} \cup \{0\}
\]
where $\N=\{1,2,3,\dots\}$ denotes the natural numbers and $0$ is the extinction state i.e. no individuals in the population.

Let  $\s_t=( \n, \x)\in \S$ be the population state at time $t$. The dynamics of $\s_t$ are determined by a set of probabilistic rules that determine the contribution of each individual in the population to the population in next time step $t+1$. Consistent with standard branching process theory, we assume that each individual in the population updates independent of all other individuals. The update rule for an individual of size $x$ is given by a probability measure $m(x,d\s)$ on the state space $\S$. Specically, the probability an individual in state $x$ contributes $\s$ individuals to the population in the next time step where $\s$ lies in a subset $A\subset \S$ equals
\[
\P[\s_1\in A| \s_0 =(1,x)]=\int_A m(x,d\s)
\]
where the left hand side reads "the probability the population state lies $A$ at time $t=1$ after initially having only one individual in state $x$." These "contributions" may correspond to an individual surviving and changing state (e.g. growing in size), or having offspring. 

The stochastic dynamics of the entire popoulation is given by a probabilistic update rule determined by the probability measures $m(x,d\s)$. Specifically, if the population state is currently $\s_t=(n_1,\dots,n_k,x_1,\dots,x_k)$, then 

1. for each of the $n_1$ individuals in state $x_1$, randomly and indepedently choose the number of replacement individuals from distribution $m(d\,\s,x_1)$.
2. repeat step 1. for the types $x_2,\dots, x_k$.
3. determine the new population state by identifying the states of all individuals and counting the total number of individuals in each of these states. 

This iterative algorithm can be used to create individual based simulations of the individual based IPM. As with any branching process, stochastic realizations of this process, with probability one, either go to extinction in finite time, or the population size asymptotically approaches infinity. 

### Probability generating functionals and extinction probabilities

As with single- and multiple-type branching processes, we can characterize the probabilistic state of the system using probability generating functionals $\Phi$ (pgfs). In the case of these individual-based IPMs, these pgfs take continuous functions $h:X\to \R$  to continuous functions of the same type. To define this pgf, we introduce the following notation. Given a continuous function $h:X\to\R$ and $\s\in \S$, let 
\[
h^\s=
\left\{
\begin{array}{cc}
\prod_{i=1}^k h(x_i)^{n_i}&\mbox{ if }\s=(n_1,\dots,n_k,x_1,\dots,x_k)\\
1 &\mbox{ if }\s=0
\end{array}
\right.
\]
Using this notation, we define the pgf by 
\[
\Psi(h)(x)=\int h^\s \,m(d\s,x).
\]

The relevance of the pgf to extinction probabilities follows from two facts. The first fact, which stems immediately from the definition, is that if $h_0$ is the zero function (i.e. $h_0(x)=0$ for all $x$), then 
\[
\Psi(h_0)(x)=\int 1 \,m(d\s,x)= \P[\s_1=0|\s_0=(1,x)]
\]
is the probability the population goes extinct in one time step given that initially it consisted of one individual is state $x$. To state the second fact, define 
\[
\Psi^t(h)=\underbrace{\Psi(\Psi(\dots \Psi}_{\mbox{$t$ times}}(h)\dots))
\]
to be the $t$-fold composition of $\Psi$ with itself. We claim that 
\[
\Psi^t(h_0)(x)=\P[\s_t=0|\s_0=(1,x)]
\]
is the population is extinct by time $t$ (or sooner) given that it initially consisted of one individual in state $x$. To see why, we argue by induction. The first fact proves that the statement holds for $t=1$. Now suppose the assertion holds for $t$, we will show it holds for $t+1$.  On the event that $\s_1=(n_1,\dots,n_k,x_1,\dots,x_k)$, extinction occurs by time $t+1$ only if each of the lineages of the $n_1+\dots +n_k$ go extinct in the next $t$ time steps. As the fates of these lineages are independent of one another, it follows that 
\[
\begin{aligned}
\P[s_{t+1}=0|\s_0=(1,x),\s_1=(n_1,\dots,n_k,x_1,\dots,x_k)]&= \prod_{i=1}^k \P[\s_t=0|\s_0=(1,x_i)]^{n_i}\\
&=\prod_{i=1}^k\left(\Psi^t(h_0)(x_i)\right)^{n_i}=\Psi_t(h_0)^{\s_1}
\end{aligned}
\]
where the second equality follows from our inductive hypothesis. By the law of total probability
\[
\P[s_{t+1}=0|\s_0=(1,x)]=\int [\Psi^t(h_0)]^\s \,m(d\s,x)
\]
which, by definition, equals $\Psi^{t+1}(0)(x)$ as claimed. 

As individuals update independent of one another, the probability of the population going extinct by time $t$ for any inititial condition $\s=(n_1,\dots,n_k,x_1,\dots,x_k)$ equals 
\[
\P[\s_t=0|\s_0=\s]=\Psi(h_0)^\s.
\]
This analytic expression allows to efficiently compute extinction probabilities by constructing a numerical approximation of the pgf and iterating it with an initial condition of a zero vector which corresponds to the numerical approximation of the zero function. 

As $\Psi_0(0)(x)\le \Psi_1(0)_1(x),\le \Psi_2(0)(x)\dots$ for any $x\in X$, there is a well defined limit corresponding to the probability of eventual extinction:
\[
q_\infty(x):=\lim_{t\to\infty} \Psi_t(0)(x).
\]
Using moment generating functionals, Harris provided a criterion for when $q_\infty(x)<1$ for all $x$ or $q_\infty(x)=1$ for all $x$. This limit theorem is based on the IPM corresponding to the expected dynamics of the individual based IPM. Namely, define the probability measure
\[
K(dy,x)=\int_{\{s=(n_1,\dots, n_k,x_1,\dots,x_k)\in \S| x_1=y\}} n_1  m(d\s,x)
\]
which defines the following IPM
\[
N_{t+1}(dy)=\int N_t(x) 
\]


### Sensitivity Formulas


## A Numerical Illustration

*I NEED TO UPDATE THE STUFF BELOW*


While the set of probabilitistic update rules can be fairly general, we focus on a relatively standard IPM set up in which there is size-dependent survival followed by size-dependent growth and reproduction. Let $s(x)$ be the probability that an individual of state $x$ survives to the next time step. Let $G(y,x)$ be the growth kernel conditioned on survival i.e. $G(y,x)dy$ is the ``infinitesmal'' probability that a surviving individual of size $x$ grows to size $y$ in the next year. Let $p_k(x)$ be the probability that an individual of size $x$ produces $k$ offspring. Let $e(y,x)dy$ be the infinitesmal probability that an offspring from a parent of size $x$ is of size $y$. We assume that the size of each offspring is independent of all other offspring and the total number of offspring. While this assumption may be reasonable for some species (e.g. many plants and sessile invertebrates), it is likely to be violated for other species (e.g. large mammals or birds). Thus, future development of these methods should account for these correlations.  

Under these assumptions, the state $\s_{t+1}$ of the population is determined by the following update rule. *For each individual in the population at time $t$,*

1. The individual survives with probability $s(x)$ where $x$ is the size of the focal individual. 
2. If the individual survives, then draw its size $y$ from the density $G(y,x)$
3. If the individual survives, randomly draw its number of offspring from the distribution $p_0(x),p_1(x),\dots$
4. For each offspring, randomly draw its size from the density $e(y,x)$

### Probability generating functionals and extinction probabilities
As with single- and multiple-type branching processes, we can characterize the probabilistic state of the system using probability generating functionals $\Phi$ (pgfs). In the case of these individual-based IPMs, these pgfs take continuous functions on $[0,1]$ to continuous function on $[0,1]$.  

To write down  $\Phi$, we need the probability generation function for the offspring distribution for an individual of size $x$: 
\[
\psi_x(z)=\sum_{k=0}^\infty p_k(x)z^k
\]
where $z$ takes on values in $[0,1]$. Recall that  $\psi_x(z)$ is called a probability generating function as it includes the probabilistic information about the offspring distribution. Specifically, 
\[
\frac{d^k}{dz^k}\psi_x(0)/k!=p_k(x)=\mbox{probability of having $k$ offspring}
\]
and
\[
\frac{d}{dz}\psi_x(1)=\sum_k k p_k(x)=\mbox{expected number of offspring}.
\]

With $\psi_x$ in hand, probability generating functional $\Phi$ is given by
\[
\Phi(h)(x)=1-s(x)+s(x)\times \int g(y,x)h(y)dy\times \psi_x\left(\int h(y)e(y,x)dy\right).
\]
This functional has the property that 


This functional has the nice property that if $\s_0=(1,x)$ then the probability of extinction by $t$ of the population is given by 
\[
E(t,x):=\Phi^t(h)(\mathbf{0})(x)
\]
where $\mathbf{0}$ denotes the zero function from $[0,1]$ to $[0,1]$ and $\Phi^t$ denotes the composition of $\Phi$ with itself $t$ times. More generally, for any initial states $\s_0=(\n,\x)$ of the popoulation, independence among individuals implies that the proability of extinction by time $t$ is given by 
\[
\prod_i E(t,x_i)^{n_i}.
\]

### Sensitivity formulas

Suppose that the IBIPM depends on some parameter $a$. Then the pgf $\Phi$ and extinction probabilities $E(t,x)$ will also depend on $a$. A simple computation reveals that the function $E(t,\cdot)$ satifies the following recursion
\[
\frac{\partial E}{\partial a}(t,\cdot)=
\frac{\partial \Phi}{\partial a}(E(t-1,\cdot))+D\Phi(E(t-1,\cdot))\frac{\partial E }{\partial a}(t-1,\cdot)
\]
where $D\Phi$ denotes the derivative of $\Phi$ 
and, consequently, can be solved for interatively. 





First, clear all the variables and load the build of the IPM which I saved in the IPMs.Rdata file. Then pull out the key functions from the IPM build which are 
$s(x)$ the size-dependent survival function, probability density $g(y,x)$ of growing from size x to y, the mean number of offspring $f(x)$ produced by an individual of size $x$ given the individual is having offspring, the probability $p_f(x)$ of having offspring at size $x$, and $os(y)$ the probability density of an offspring being size $y$ at the next census.

```{r}
rm (list = ls ())
load("IPMs.Rdata")
g=function(y,x)Growth(y,x)
s=function(x)Survival(x)
f=function(x)f.fecundity(x)*establishment.prob
pf=function(x)1
os=function(y)dnorm(y,mean=recruit.size.mean,sd=recruit.size.sd)
```
Now comes the important new piece. Need to define the probability generating function associated with the fecundity function. Since we used the Poisson family for getting the fecundity function, I assume that the number of offspring is Poisson distributed with mean $f(x)$ for an individual of size $x$. Hence, the probability generating function (with size $x$ treated like a parameter) is given by  

```{r}
phi=function(x,s)exp(f(x)*(s-1))
```

Using all of these pieces, we can define the probability generating functional for individual based IPM by 
\[\Psi(h)(x)=\left(1-p_f(x)+p_f(x)\phi\left(x,\int_a^b os(y) h(y) dy\right)\right)\left(1-s(x)
						+s(x)\int_a^b g(y,x)h(y)dy\right)\]
where $h$ is a continuous function from $[a,b]$ to $[0,1]$.

To create this probability generating functional numerically, I use the grid on x values xs which go from the minimum size $\alpha$ to the maximum size $\beta$ with n bins. 

```{r}
n=100
xs=seq(alpha,beta,length=n)
dx=xs[2]-xs[1]
```

Now create the survival, probability of having offspring, and offspring size vectors. 

```{r}
s.mat=s(xs)
pf.mat=pf(xs)
os.mat=os(xs)*dx
```
We need to make sure that the o.mat integrates to one i.e. there is no eviction. Eviction creates a major problem for the probability generating functionals. 

```{r}
sum(os.mat)
```
As this sum is pretty close to one, I just renormalize it. 

```{r}
os.mat=os.mat/sum(os.mat)
sum(os.mat)
```

Next need to define the growth matrix and deal with the eviction issue. In this case, I am treating eviction as mortality as this is consistent with what was being done in the Nicols et al. paper. The new growth model is s.evict*g.evict which agrees with the orginal growth kernel g but has a column stochastic matrix g.evict. Note: Most of this eviction is occuring at negative sizes. 

```{r}
g.mat=outer(xs,xs,g)*dx
q.evict=1-colSums(g.mat) 
g.evict=g.mat%*%diag(1/(1-q.evict))
s.evict=(1-q.evict)*s.mat
```

Now, we are ready to create the discretized probability generation function which takes vectors of length n (i.e. the discretized function h) and returns vectors of length n. Note: I need to take the transpose to take integrals down columns.

```{r}
Psi=function(h){
	h.new=(s.evict*(t(g.evict)%*%h)+1-s.evict)*(1-pf(xs)+pf(xs)*phi(xs,os.mat%*%h))
	return(h.new)
}
```

We can use this functional to determine how the probability of extinction varies in time. In particular, iterating the $h=0$ function for $t$ time steps yields $\Psi^t(h)$ where $\Psi^t(h)(x)$ is the probability of the population going extinct by year $t$ given initially there was one individual of size $x$ in the population. The following function iterates the discretized $h=0$ function until time $T$ and returns a matrix of dimension $n\times (T+1)$ whose $t-1$-th column correspondds to $\Psi^t(h)$.
 
```{r}
Psi.iterate=function(T){
	hs=matrix(0,T+1,n)
	for(t in 1:T){
		hs[t+1,]=Psi(hs[t,])
		}
	return(hs)
}
```

OK. Lets try this function out and plot the probabilities of extinction as functions of time for several sizes. 

```{r}
T=15
out=Psi.iterate(T)
sizes=seq(1,n,length=50)
matplot(0:T,out[,sizes],type='l',lwd=2,lty=1,bty="n",col=grey(seq(0.1,0.9,length=length(sizes))),xlab="time",ylab="probability of extinction by time t")
```

Can also plot exinction by time 10 as a function of the initial size of the "founding individaul"

```{r}
plot(out[11,],type="l",lwd=3,xlab="size",ylab="probability of extinction by time 10")
```

To scale things up to an entire population, we can introduce the following function with two inputs: N vector of length $n$ which specifies the initial number of individuals in each size bin, and $T$ the length of the time to look things over. The function returns a vector of length $T+1$ corresponding to the probabilities of extinction by time $t$ or sooner. 

```{r}
PVA=function(N,T){
	hs=Psi.iterate(T)
	output=numeric(T+1)
	for(t in 1:(T+1)){
		output[t]=prod(hs[t,]^N)
	}
	return(output)
}
```

Lets try this out by considering a population of 100 individuals of the smallest size.

```{r}
T=25
N=rep(0,n)
N[1]=100
out=PVA(N,T)
plot(0:T,out,typ="l",lwd=3,bty="n",xlab="time t",ylab="probability of extinction by year t")
abline(h=0.01)
```

Clearly these probabilities asymptote. The following function computes this asymptotic probability of extinction by iterating toward the steady state. 

```{r}
AE=function(tol=0.0001){
  h=rep(0,n)
  diff=1
  while(diff>tol){
    h.old=h
    h=Psi(h)
    diff=sum(abs(h.old-h))
  }
  return(h)
}
```
Lets try this out. 

```{r}
h=AE()
plot(xs,h,type="l",xlab="size",ylab="asymptotic extinction probability",lwd=3)
```

Can see that the asymptotic exinction probability tends to decrease with the size of the founding individual but at large sizes begins increasing again. This latter trend is due to the eviction in the model at large sizes.
