---
title: 'Individual-based Integral Projection Models: The role of size-structure on
  extinction risk and establishment success'
author: "Sebastian Schreiber and Noam Ross"
date: "April 8, 2015"
output: html_document
---
$\def\N{\mathbb N}$
$\def\R{\mathbb R}$
$\def\P{\mathbb P}$
$\def\N{\mathbb N}$
$\def\S{\mathcal S}$
$\def\n{\mathbf n}$
$\def\s{\mathbb s}$
$\def\x{\mathbf x}$
$\def\y{\mathbf y}$
$\def\o{\%*\%}$

##Introduction

Computations of extinction probabilities or likelihoods of establishment success lie on opposing sides of a theoretician's coin and are used to address questions in conservation biology, restoration ecology, biological invasions and population genetics. Risks of extinction or establishment failure stem from populations consisting of a finite number of individuals each of which faces a non-zero risk of mortality on any given day. These risks are shaped, in part, by the size and composition of a population in which individuals may differ in age, size, geographical location, or other important characteristics influencing demography. When population structure is finite-dimensional (i.e. a finite number of age classes, stages, geographical locations), multi-type branching processes serve as the stochastic counterpart of matrix models (A & Ney 19XX, Caswell 2001). These models allow one to examine how extinction risk or establishment success depend on the population size and its distribution across a finite number of states (e.g. ages, size classes, patches). 

When collecting demographic data, one commonly makes continuous measurements (e.g. mass, length) about individuals. Consequently, since their introduction in the early 2000s, there has been a surge of interest in integral projection models (IPMs) that account for continuous population structure (REFs). As IPMs correspond to infinite-dimensional matrix models which are nummerically approximated by finite-dimensional matrix models, many of the standard demographic concepts and methods for matrix models (e.g. stable state distributions, reproductive values, life table response experiements) extend to IPMs (REFs).  

Here, we describe individual-based counterparts of IPMs by making use of the theory of continuous-state branching processes (Harris XX). We present new analytical methods for computing extinction probabilities for these individual-based IPMs. These general methods allow one to examine how extinction probabilities depend on any mixture of  continuous and discrete population structure. We also introduce methods for examining the sensitivity of the extinction probabilities to model parameters. To illustrate the implementation of these methods, we introduce and analyze an individual-based IPM of XXX, which is a stochastic counterpart of the IPM developed by XXX. 



##The General Models and Methods

### The Individual Based IPM
We consider an individual-based model where the set of all possible individual states (e.g. age, size, etc.) lies in a compact metric space $X$. For example, for the standard size-structured IPM, $X=[a,b]$ corresponds to the range of sizes measured in the field. For models with a mixture of age and size stucture, $X$ could be given by $\{1,\dots,T\}\times [a,b]$ where $T$ corresponds to the maximal age of an individual.

As we consider finite populations, the state of the population at any point of the time is characterized by the different sizes $(x_1,x_2,\dots, x_k)$  of individuals within the population and how many individuals there are of each these sizes $(n_1,n_2,\dots, n_k)$. Specifically, if there $n_1$ individuals of size $x_1$, $n_2$ individuals of size $x_2$,..., $n_k$ individuals of size $x_k$, then the state of the population is given by $(n_1,n_2,\dots, n_k; x_1,x_2,\dots, x_k)$. Following Harris (1963), the set of all possible population states is
\[
\S= \{(\n,\x)=(n_1,\dots, n_k, x_1\dots , x_k): k,n_i\in \N, x_i \in X\} \cup \{0\}
\]
where $\N=\{1,2,3,\dots\}$ denotes the natural numbers and $0$ is the extinction state corresponding to no individuals in the population.

Let  $\s_t=( \n, \x)\in \S$ be the population state at time $t$. The dynamics of $\s_t$ are determined by a set of probabilistic rules that determine the contribution of each individual in the population to the population in next time step $t+1$. These "contributions" may correspond to an individual surviving and changing state (e.g. growing in size), or having offspring.  Consistent with standard branching process theory, each individual in the population updates independent of all other individuals. The update rule for an individual of size $x$ is given by a probability measure $m(x,d\s)$ on the state space $\S$. Specically, the probability an individual in state $x$ contributes $\s$ individuals to the population in the next time step where $\s$ lies in a subset $A\subset \S$ equals
\[
\P[\s_1\in A| \s_0 =(1,x)]=\int_A m(x,d\s)
\]
where the left hand side reads "the probability the population state lies $A$ at time $t=1$ after initially having only one individual in state $x$." 

If the population state is currently $\s_t=(n_1,\dots,n_k,x_1,\dots,x_k)$, then the state $\s_{t+1}$ is determined as follows:

1. for each of the $n_1$ individuals in state $x_1$, randomly and indepedently choose the number of replacement individuals from distribution $m(x_1,d\,\s)$.
2. repeat step 1. for the types $x_2,\dots, x_k$.
3. determine the new population state by identifying the states of all individuals and counting the total number of individuals in each of these states. 

This iterative algorithm can be used to create individual based simulations of the individual based IPM. As with any branching process, stochastic realizations of this process, with probability one, either go to extinction in finite time, or the population size grows without bound. 

### Probability generating functionals and extinction probabilities

As with multiple-type branching processes, we can characterize the probabilistic state of the system using probability generating functionals $\Psi$ (pgfs). Our approach differs from Harris (1963) who used moment generating functionals instead. Unlike mgfs, the pgfs allow us to compute how extinction probabilties change in time as well as compute asymptotic extinction probabilities. 

To define the pgf $\Psi$, we introduce the following notation: given a continuous function $h:X\to\R$ and $\s\in \S$, let 
\[
h^\s=
\left\{
\begin{array}{cc}
\prod_{i=1}^k h(x_i)^{n_i}&\mbox{ if }\s=(n_1,\dots,n_k,x_1,\dots,x_k)\\
1 &\mbox{ if }\s=0.
\end{array}
\right.
\]
We define the pgf by 
\[
\Psi(h)(x)=\int h^\s \,m(x,d\s).\tag{*}
\]
In words, $\Psi(h)(x)$ corresponds to the expected value of $h^\s$ due to the contributions of individuals in the next time step from an individual in state $x$. This requires integrating over all possible populations states in the next time step.  

The utility of $\Psi$ for computing extinction probabilities follows from two facts. The first fact stems immediately from the definition: if $h_0$ is the zero function (i.e. $h_0(x)=0$ for all $x$), then 
\[
\Psi(h_0)(x)=\int 1 \,m(d\s,x)= \P[\s_1=0|\s_0=(1,x)]
\]
is the probability the population goes extinct in one time step given that initially it consisted of one individual is state $x$. For the second fact, we define 
\[
\Psi^t(h)=\underbrace{\Psi(\Psi(\dots \Psi}_{\mbox{$t$ times}}(h)\dots))
\]
to be the $t$-fold composition of $\Psi$ with itself. We claim that 
\[
\Psi^t(h_0)(x)=\P[\s_t=0|\s_0=(1,x)]=:E_t(x)\tag{**}
\]
is the probability of extinction by time $t$ given that the population initially consisted of one individual in state $x$. To see why this fact is true, we argue by induction. The first fact proves that the statement holds for $t=1$. Now suppose the assertion holds for $t$, we will show it holds for $t+1$.  On the event that $\s_1=(n_1,\dots,n_k,x_1,\dots,x_k)$ is the population state at time $1$, extinction occurs by time $t+1$ only if each of the lineages of the $n_1+\dots +n_k$ individuals go extinct in the next $t$ time steps. As the fates of these lineages are independent of one another, it follows that 
\[
\begin{aligned}
\P[s_{t+1}=0|\s_0=(1,x),\s_1=(n_1,\dots,n_k,x_1,\dots,x_k)]&= \prod_{i=1}^k \P[\s_t=0|\s_0=(1,x_i)]^{n_i}\\
&=\prod_{i=1}^k\left(\Psi^t(h_0)(x_i)\right)^{n_i}=\Psi^t(h_0)^{\s_1}
\end{aligned}
\]
where the second equality follows from our inductive hypothesis. By the law of total probability
\[
\P[s_{t+1}=0|\s_0=(1,x)]=\int [\Psi^t(h_0)]^\s \,m(d\s,x)
\]
which, by definition, equals $\Psi^{t+1}(0)(x)$ as claimed. 

Hence, equation (**) can be used to compute extinction probabilities iteratively. Furthremore, as individuals update independent of one another, the probability of the population going extinct by time $t$ for any initial condition $\s=(n_1,\dots,n_k,x_1,\dots,x_k)$ equals 
\[
\P[\s_t=0|\s_0=\s]=E_t^\s.
\]
These analytic expressions allow us to efficiently compute extinction probabilities by constructing a numerical approximation of the pgf and iterating it with an initial condition of a zero vector which corresponds to the numerical approximation of the zero function. 

As $E_0(x)\le E_1(x),\le E_2(x)\dots$ for any $x\in X$ and $E_t(x)\le 1$ for all $t$, there is a well defined limit corresponding to the probability of eventual extinction:
\[
E_\infty(x):=\lim_{t\to\infty} E_t(x).
\]
Using moment generating functionals and under suitable technical hypotheses, Harris showed that $E_\infty(x)<1$ for all $x$ if the dominant eigenvalue of the mean-field IPM is greater than one, and $E_\infty(x)=1$ for all $x$ otherwise.



### Sensitivity Formulas

To examine the sensitivity of the extinction probability $E_t(x)$ to a parameter $a$ in an individual-based IPM such as the slope or intercept of the logistic regression for a survival function, the chain rule implies that 
\[
\frac{d}{da}E_t(x)=\left(\prod_{s=0}^{t-1} \frac{d\Psi}{da}(E_s)\right)(x)
\]
where the product refers to composition of the linear operators given by taking the derivative of the pgf with respect to the parameter $a$. From these sensitivities, we can compute the elasticities of these extinction probabilities to $a$ which describe proportional changes in $E_t(x)$ due to proportional changes in $a$:
\[
 \frac{a}{E_t(x)}\left(\prod_{s=0}^{t-1} \frac{d\Psi}{da}(E_s)\right)(x).
\]

## An Illustration with an Endangered Plant Species

To illustrate how these general methods can be applied to a data set, we develop an individual-based IPM based on **add details about the endangered plant species and the work of Nicols et al**. 

For the mean-field IPM, Nicols et al. had a model of the form 
\[
N_{t+1}(x)=\int_a^b s(y)G(y,x)+e(x)f(y)N_t(y)\,dy
\]
where $s(x)$ is the probability of surviving to the next year for individuals of size $y$, $G(y,x)dy$ is the infinitesimal probability that a surviving individual of size $x$ is size $y$ in the next year, $f(x)$ is the mean number of offspring produced by an individual of size $x$, and $e(x)$ is the infinitesimal probability of an offspring is of size $x$ at the time of the annual census. 

The following code loads the functions for the mean field IPM: 
```{r}
rm (list = ls ())
load("IPMs.Rdata")
g=function(y,x)Growth(y,x)
s=function(x)Survival(x)
f=function(x)f.fecundity(x)*establishment.prob
pf=function(x)1
e=function(y)dnorm(y,mean=recruit.size.mean,sd=recruit.size.sd) 
```

The kernels $s(x)$, $G(y,x)$, and $e(x)$ provide us with all the information the individual-based IPMs requires for probabilistic updating individuals for survival, growth, and size at first census. The fecundity kernel $f(y)$, however, only specifies the mean number of offspring produced, but for an individual-based IPM, we need the distribution of the number of offspring produced by an individual. Fortunately, when using GLMs to model fecundity data, we are often getting this extra information for free. For example, one might use the Poisson family in a GLM to model the fecundity data.  In which case, the mean $f(x)$ also specifies the distribution. Alternatively, one might use multi-parameter distributions such as the negative binomial or mixtures of several distributions.

Consistent with the analysis of Nicols et al, we assume this distribution for an individual of size $y$ is Poisson distributed with mean $f(y)$. Recall that pgf for a Poisson distribution with mean $\lambda$ is $\phi(s)=\exp(\lambda(s-1))$ where $s$ is a dummy variable. Hence, the pgf for the number of offspring of an individual of size $x$ is given by 
```{r}
phi=function(x,s)exp(f(x)*(s-1))
```
We will need this pgf to construct the pgf $\Psi$ of the individual-based IPM. 


### Deriving the pgf $\Psi$

To define $\Psi$, notice that the contributions of an individual of size $x$ to the population in the next time step involves the sum of two independent random variables. There is the contribution due to survival and growth, and the contribution due to reproduction. We can define a pgf for each of these processes separately. For the survival and growth pgf $\Psi_g$, $\Psi_g(h)(x)$ corresponds to integrating $h^\s$ over all contributes $\s$ due to an individual of size $x$ surviving and growing. These possible states are of two types: $\s=\emptyset$ corresponding to an individual dying and $\s=(1,y)$ corresponding to an individual surviving and growing to size $y$. The first event occurs with probability $1-s(x)$ and the infinitesimal probability of the second event is $s(x)G(y,x)dy$. As $h^\s=1$ on the event of death and $h^\s=h(y)$ on the event $\s=(1,y)$, integrating over all possible contributions due to survival and growth yields 
\[
\Psi_{g}(h)(x)=(1-s(x))\times 1 + s(x)\int h(y)G(y,x)dy.
\]
For the pgf $\Psi_f$ corresponding to contributions due to fecundity, $\Psi_f(h)(x)$ is given by integrating $h^\s$ over all possible states $\s$ corresponding to the offspring produced by an individual of size $x$. With this two pgfs, we can use a fundamental property of pgfs: the pgf for a sum of independent random variables is the product of the pgfs of these random variables. Hence, our pgf $\Psi$ for the individual-based IPM will be given by 
\[
\Psi=\Psi_g \times \Psi_f
\]

But what about $\Psi_f$? To write this down, we can take advantage of the fact that the reproductive contribution of an individual of size $x$ corresponds to a Poisson number of offspring with mean $f(x)$ whose sizes are drawn *independently* from the same offspring distribution. To make use of this observation, we can condition on the number of offspring an individual has. Lets say our individual of size $x$ has $N$ offspring. Then its contribution is a sum of $N$ independent random variables taking values of the form $\s=(1,y)$. The pgf $\Psi_{kid}$ associated with one of these contributions is given by 
\[
\Psi_{kid}(h)(x)=\int h(y)e(y)dy
\]
as $h^{{1,y}}(x)=h(y)$ and $e(y)dy$ is the infinitesimal probability of an offspring being of size $y$. Since there are $N$ kids independent of one another, the pgf for this sum is given by the product of the pgfs:
\[
\Psi_{N\,kids}(h)(x)=\left(\int h(y)e(y)dy\right)^N.
\]
Finally, to get $\Psi_f$, we can sum over all possible number of kids weighted by their probabilities:
\[
\Psi_f(h)(x)=\sum \frac{\exp(-f(x))f(x)^N}{N!} \Psi_{kid}(h)(x)^N
\]
but this corresponds simply to 
\[
\Psi_f(h)(x)=\phi(x, \Psi_{kid}(h)(x))=\phi(x,\int h(y)e(y)dy)
\]
Thus we have
\[
\Psi(h)(x)=\left((1-s(x))\times 1 + s(x)\int h(y)G(y,x)dy\right)\phi(x,\int h(y)e(y)dy)
\]

### Numerically Implementing and Utilizing $\Psi$

To create this probability generating functional numerically, we discretize the size interval $[\alpha,\beta]$ using $n$ equal sized intervals of width $dx=(\beta-\alha)/n$. **NEED TO MAKE SURE EVERYTHING IS CONSISTENT WITH MIDPOINT RULE**

```{r}
n=100
xs=seq(alpha,2*beta,length=n)
dx=xs[2]-xs[1]
```

Now create the survival and offspring size vectors. For illustrative purposes, we reduced survivorship of all individuals by 10\%. 

```{r}
s.mat=s(xs)*0.9
e.mat=e(xs)*dx
```
We need to make sure that the e.mat integrates to one i.e. there is no eviction. Eviction creates a major problem for the probability generating functionals. 

```{r}
sum(e.mat)
```
As this sum is pretty close to one, we just renormalize it. 

```{r}
e.mat=e.mat/sum(e.mat)
```

Next, we define the growth matrix and deal with the eviction issue. We choose to am treat eviction as mortality to be consistent with Nicols et al. The new growth model is s.evict*g.evict which agrees with the original growth kernel g but has a column stochastic matrix g.evict. Note: Most of this eviction is occurring at negative sizes. 

```{r}
g.mat=outer(xs,xs,g)*dx
q.evict=1-colSums(g.mat) 
g.evict=g.mat%*%diag(1/(1-q.evict))
s.evict=(1-q.evict)*s.mat
```

We can create the discretized pgf $\Psi$ which takes vectors of length $n$ (i.e. the discretized function $h$) and returns vectors of length $n$. 

```{r}
Psi=function(h){
	h.new=(s.evict*(t(g.evict)%*%h)+1-s.evict)*(1-pf(xs)+phi(xs,e.mat%*%h))
	return(h.new)
}
```

To compute how the probability of extinction varies in time, the following function iterates the zero vector (the discretized $h_0$ function) until time $T$ and returns a matrix of dimension $n\times (T+1)$ whose $t-1$-th column corresponds to $\Psi^t(h_0)$. 
 
```{r}
Psi.iterate=function(T){
	hs=matrix(0,T+1,n)
	for(t in 1:T){
		hs[t+1,]=Psi(hs[t,])
		}
	return(hs)
}
```

Lets try this function out and plot the probabilities of extinction as functions of size for a subset of years. 

```{r}
T=100
out=Psi.iterate(T)
times=c(1,5,10,25,50,100)+1
matplot(xs,t(out[times,]),xlim=c(alpha,beta),type="l",lty=1,xlab="size",ylab="extinction probabilities",bty="n")
legend("topright",as.character(times-1),bty="n")
```

Intuitively, this figure illustrates that (i) extinction probability decreases if the size of one individual initially in the population is larger, and (ii) extinction probabilities increase over time. Furthermore, this figure illustrates that $E_t(x)$ are approaching limiting extinction probabilities $E_{\infy}(x)$ which are less than one. This stems from the fact that the dominant eigenvalue of the mean-field IPM is greater than one. The following function computes this asymptotic probability of extinction by iterating until a tolerance condition is met.  

```{r}
AE=function(tol=0.0001){
  h=rep(0,n)
  diff=1
  while(diff>tol){
    h.old=h
    h=Psi(h)
    diff=sum(abs(h.old-h))
  }
  return(h)
}
```

Using this function yields the following plot of $E_\infty(x)$

```{r}
h=AE()
plot(xs,h,type="l",xlab="size",ylab="asymptotic extinction probability",lwd=3,xlim=c(alpha,beta))
```


To scale things up to an entire population, we can introduce the following function with two inputs: N vector of length $n$ which specifies the initial number of individuals in each size bin, and $T$ the length of the time to look things over. The function returns a vector of length $T+1$ corresponding to the probabilities of extinction by time $t$ or sooner. 

```{r}
PVA=function(N,T){
	hs=Psi.iterate(T)
	output=numeric(T+1)
	for(t in 1:(T+1)){
		output[t]=prod(hs[t,]^N)
	}
	return(output)
}
```

Lets try this out by considering a population of 100 individuals of the smallest size.

```{r}
T=25
N=rep(0,n)
N[1]=100
out=PVA(N,T)
plot(0:T,out,typ="l",lwd=3,bty="n",xlab="time t",ylab="probability of extinction by year t")
abline(h=0.01)
```

