---
title: Re-implementing the IPM script using IPMpack
author: Noam Ross
date: February 4, 2015
output:
  html_document:
    keep_md: true
---

Read in the data

```{r setup}
library(IPMpack)
library(knitr)
data = read.csv("../../data/IPM-data.csv")
opts_chunk$set(cache=TRUE)
```

First, make growth and survival objects

```{r}
growth = makeGrowthObj(dataf = data, Formula = sizeNext~size)
surv = makeSurvObj(dataf = data, Formula = surv ~ size)
```

Each of these objects have a slot called "fit" which contains the full
glm model, so we can access them and plot them:

```{r}
alpha=min(data$size[which(data$size>0)],data$sizeNext[which(data$sizeNext>0)])
beta=max(data$size[which(data$size>0)],data$sizeNext[which(data$sizeNext>0)])
xs=seq(alpha,beta,length=101)
Survival=function(x)predict(surv@fit,data.frame('size'=x),type='response')
plot(data$size,jitter(data$surv,factor=0.1),pch=21,bg=rgb(1,0,0,0.5))
lines(xs,Survival(xs),lwd=4,col="blue")
```

OK, so far, not that useful.  Wait!  Just discovered built-in plotting
functions:

```{r}
picGrow(data, growth)
picSurv(data, surv, ncuts=101)
```

Sebastian's plot is prettier.

Now the fecundity object, which is a little different:

```{r}
fec = makeFecObj(dataf = data, Formula = fec.seed ~ size, Family = 'poisson',
								 fecConstants = data.frame(sum(is.na(data$size))/
								 														sum(data$fec.seed,na.rm=TRUE)),
								 meanOffspringSize = mean(data$sizeNext[is.na(data$size)]),
								 sdOffspringSize = sd(data$sizeNext[is.na(data$size)]))
```

For whatever reason the fit object in `fec` is `fec@fitFec[[1]]`.  There's no
plot method for this.  Here's your plot of the same.  It looks like the same fit.

```{r}
plot(data$size,data$fec.seed,pch=21,bg=rgb(1,0,0,0.5),
		 bty="n",xlab="size",ylab="fecundity")
lines(xs,predict(fec@fitFec[[1]], data.frame('size'=xs),type="response"),lwd=2)
```

So now we can build our matrix.  In **IPMpack**, we nest these, starting with
just the growth-by-survivial (P) matrix.  Looks like there is a midpoint
integration option.  Then we can extact various diagnostics:

```{r}
Pmat = makeIPMPmatrix(growObj = growth, survObj = surv, minSize=alpha,
											maxSize=beta, nBigMatrix = 101,
											integrateType = "midpoint")
diagnosticsPmatrix(Pmatrix = Pmat, growObj = growth, survObj = surv, dff = data)
meanLifeExpect(Pmat)
passageTime(mean(data$size, na.rm = TRUE), Pmat)
```

Now let's add fecundity.

```{r}
Fmat = makeIPMFmatrix(fec, minSize = alpha, maxSize = beta, nBigMatrix=101,
											survObj = surv, growObj = growth)
```

We can calculate the stable age distribution and lambda:

```{r}
calcs = largeMatrixCalc(Pmat, Fmat)
calcs$lam
plot(as.vector(calcs$stableDist))
```

And plot a contour map:

```{r}
image(Pmat + Fmat)
contour(Pmat + Fmat, add=TRUE)
```

I can handle the zero-inflation in the fecundity data, too, by creating a
fecundity function that explictly deals with flowering. 

```{r}
data2= data
data2$fec.seed[data$fec.flower == 0] = NA
fec2 = makeFecObj(dataf = data2, Formula = list(fec.flower ~ size, fec.seed ~ size),
								 Family = c('binomial', 'poisson'),
								 fecConstants = data.frame(1, sum(is.na(data$size))/
								 														sum(data$fec.seed,na.rm=TRUE)),
								 Transform = c("none", "-1"),
								 meanOffspringSize = mean(data$sizeNext[is.na(data$size)]),
								 sdOffspringSize = sd(data$sizeNext[is.na(data$size)]))
Fmat2 = makeIPMFmatrix(fec2, minSize = alpha, maxSize = beta, nBigMatrix=101,
											survObj = surv, growObj = growth)
```

Here are the flowering probability and seed production plots

```{r}
plot(data$size,jitter(data$fec.flower,factor=0.1),pch=21,bg=rgb(1,0,0,0.5),
		 bty="n",xlab="size",ylab="flowering probability")
lines(xs, predict(fec2@fitFec[[1]], data.frame('size'=xs),type="response"),lwd=2)


plot(data$size,data$fec.seed,pch=21,bg=rgb(1,0,0,0.5),
		 bty="n",xlab="size",ylab="seed fecundity")
lines(xs,predict(fec2@fitFec[[2]], data.frame('size'=xs),type="response"),lwd=2)
```

And we can compare the old and new fecundity matrices:

```{r}
par(mfrow=c(1,2), mar=c(1,1,3,1))
persp(Fmat[seq(1,101,length.out=50), seq(1,101,length.out=50)], main="Poisson Only", zlim=c(0, 0.1), theta=20)
persp(Fmat2[seq(1,101,length.out=50), seq(1,101,length.out=50)], main="Binomial Flowering\nPoisson Seeds", zlim=c(0, 0.1), theta=20)
par(mfrow=c(1,1))
```

With this new fecundity matrix here are our outcomes:

```{r}
calcs2 = largeMatrixCalc(Pmat, Fmat2)
calcs2$lam
plot(as.vector(calcs2$stableDist))
```


OK, initial conclusions:

1.  This package pretty much does what we expect it to.
2.  I HATE it's syntax and API.


